A White Noise Approach to Evolutionary Ecology

### 2020-04-02 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
fig4Updated!

In real populations individuals are born and potentially reproduce before they ultimately die. These three events provide the basic ingredients of a branching process. Mathematical investigations of such processes have a relatively deep history [@Kendall1966]. The most simple branching process, known as the Galton-Watson process, describes the number of individuals alive at a given time $t\geq0$ as a non-negative integer [@Kimmel2015]. Feller [-@feller1951] introduced a formal method to approximate branching processes with diffusion processes which are continuous in state (i.e., population size is approximated as a continuous quantity). Since diffusion processes possess greater analytical tractability than branching processes, Feller's method, known as the diffusion limit, has acquired immense popularity particularly in the field of mathematical population genetics [@Ewens2004]. For over the past half of a century a great deal of accomplishments have been achieved in formalizing the diffusion limits of branching processes that describe populations of individuals occuring in some continuous space [@Watanabe1968; @Dawson1975; @Perkins1992; @Mlard1993; @edwinarendperkins1995; @zeng1998absolute; @Bertoin2003; @Etheridge2008; @Barton2019]. This space can represent geographic space or, relevant to our context, phenotypic space. In the following subsection, we describe the BBM process, which is a particularly important branching process with spatial stucture. This process has been very useful in the study of SPDE due to its simplifying assumption that individuals do not interact. However, this assumption imposes an unfortunate restriction by precluding the modelling of ecological interactions. We therefore follow our discussion of BBM with a review of a few important results on spatially structured branching processes that account for interactions.

@@

A convergence theorem for the diffusion limit of a generalization of BBM was established by Watanabe [-@Watanabe1968]. Dawson [-@Dawson1975] suggested that, for spatial dimension $d=1$, this diffusion limit should admit a density process that satisfies a SPDE. Konna and Shiga [-@Konno1988] and Reimers [-@Reimers1989] independently proved Dawson's suggestion was indeed correct. The diffusion limit of this more general branching process (in arbitrary spatial dimension) is referred to as a Dawson-Watanabe superprocess [@alisonetheridge2000]. Conditioning a Dawson-Watanabe superprocess to have constant mass returns a Fleming-Viot process [@Etheridge1991; @Perkins1991] which has been popular in studies of spatial population genetics. In particular, an extension of Fleming-Viot processes, known as $\Lambda$-Fleming-Viot processes, were introduced by Bertoin and Le Gall [-@Bertoin2003] and coined by Etheridge [-@Etheridge2008] where it was used to resolve some technical challenges in modelling isolation by distance [@Felsenstein1975; see also @Barton2013; and @Barton2019]. Although this provides an impressive list of accomplishments, the Dawson-Watanabe superprocess falls short of our needs. In particular this process assumes individuals do not interact and thus precludes its ability to model ecological interactions. However, this concern has been addressed, leading to constuctions of superprocesses that account for interactions among individuals. In the next subsection we summarize the main results in this area and introduce the SPDE that provides the basis for our approach to theoretical evolutionary ecology.

@@

With solutions to SPDE (\ref{SPDE}) well defined for any growth rate bounded above, we can calculate the total mass process $N(t)$ using the mild solution of (\ref{SPDE}) with $f(x)\equiv1\in C_b^2(\mathbb{R})$ (the symbol "$\equiv$" means equal to for every $x$). That is,
\begin{multline}
N(t)=N(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)\cdot1+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}1\right)+1\sqrt{V\nu(x,s)}\dot W(x,s)dsdx \\
=N(0)+\int_0^t\bar m(s)N(s)dt+\int_0^t\sqrt{VN(s)}d\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)}),
\end{multline}

where
\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}
and
\begin{equation}
\int_0^td\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)})=\int_0^t\int_\mathbb{R}\frac{\sqrt{\nu(x,s)}}{\sqrt{\int_\mathbb{R}\nu(x,s)dx}}\dot W(x,s)dxds.
\end{equation}
Setting $W_1(t)=\hat{\mathbf{W}}_t(\sqrt{\nu(x,t)})$, we can use traditional stochastic differential notation to write
\begin{equation}
dN=\bar mNdt+\sqrt{VN}dW_1.
\end{equation}
To find the associated SDE for $\bar x(t)$ and $\sigma^2(t)$, we want to repeat the same approach for $f(x)=x,x^2$ and apply It&ocirc;'s lemma. However, for these cases $f\notin C^2_b(\mathbb{R})$ since $f$ will not be bounded. However, if we can show that $\int_\mathbb{R}(|x|+x^2+x^4)\nu(x,t)dx<+\infty$ for all $t>0$ given this condition is satisfied by $\nu(x,0)$, then we can apply the mild solution of (\ref{SPDE}) to derive SDE for $\bar x(t)$ and $\sigma^2(t)$. To illustrate, let us suppose this is the case. Setting $\tilde x(t)=\int_\mathbb{R}x\nu(x,t)dx$, we have
\begin{equation}
\tilde x(t)=\tilde x(0)+\int_0^t\int_\mathbb{R}\nu(x,s)m(\nu,x)x+x\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Similarly, setting $\tilde{\tilde\sigma}^2(t)=\int_\mathbb{R}x^2\nu(x,t)dx$, we have
\begin{equation}
\tilde{\tilde\sigma}^2(t)=\tilde{\tilde\sigma}^2(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)x^2+\mu\right)+x^2\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Since $\bar x(t)=\tilde x(t)/N(t)$ and $\sigma^2(t)=\tilde{\tilde\sigma}^2(t)/N(t)-\bar x^2(t)$, we can use It&ocirc;'s lemma to derive SDE for $\bar x(t)$ and $\sigma^2(t)$, which we perform in SM \S\ref{SDE_DERIV}. We do not make any general assertions about the existence or uniqueness of $\bar x(t)$ or $\sigma^2(t)$.

>>>>

In real populations individuals are born and potentially reproduce before they ultimately die. These three events provide the basic ingredients of a branching process. Mathematical investigations of such processes have a relatively deep history [@Kendall1966]. The most simple branching process, known as the Galton-Watson process, describes the number of individuals alive at a given time $t\geq0$ as a non-negative integer [@Kimmel2015]. Feller [-@feller1951] introduced a formal method to approximate branching processes with diffusion processes which are continuous in state (i.e., population size is approximated as a continuous quantity). Since diffusion processes possess greater analytical tractability than branching processes, Feller's method, known as the diffusion limit, has acquired immense popularity particularly in the field of mathematical population genetics [@Ewens2004]. For over the past half of a century a great deal of accomplishments have been achieved in formalizing the diffusion limits of branching processes that describe populations of individuals occuring in some continuous space [@Watanabe1968; @Dawson1975; @Dawson1978; @Perkins1992; @Mlard1993; @edwinarendperkins1995; @zeng1998absolute; @Bertoin2003; @Etheridge2008; @Barton2019]. This space can represent geographic space or, relevant to our context, phenotypic space. In the following subsection, we describe the BBM process, which is a particularly important branching process with spatial stucture. This process has been very useful in the study of SPDE due to its simplifying assumption that individuals do not interact. However, this assumption imposes an unfortunate restriction by precluding the modelling of ecological interactions. We therefore follow our discussion of BBM with a review of a few important results on spatially structured branching processes that account for interactions.

@@

A convergence theorem for the diffusion limit of a generalization of BBM was established by Watanabe [-@Watanabe1968]. Dawson [-@Dawson1975] suggested that, for spatial dimension $d=1$, this diffusion limit should admit a density process that satisfies a SPDE. Konna and Shiga [-@Konno1988] and Reimers [-@Reimers1989] independently proved Dawson's suggestion was indeed correct. The diffusion limit of this more general branching process (in arbitrary spatial dimension) is referred to as a Dawson-Watanabe superprocess [@alisonetheridge2000]. Conditioning a Dawson-Watanabe superprocess to have constant mass returns a Fleming-Viot process [@Etheridge1991; @Perkins1991] which has been popular in studies of spatial population genetics. In particular, an extension of the Fleming-Viot process, known as the $\Lambda$-Fleming-Viot process, was introduced by Bertoin and Le Gall [-@Bertoin2003] and coined by Etheridge [-@Etheridge2008] where it was used to resolve some technical challenges in modelling isolation by distance [@Felsenstein1975; see also @Barton2013; and @Barton2019]. Although this provides an impressive list of accomplishments, the Dawson-Watanabe superprocess falls short of our needs. In particular this process assumes individuals do not interact and thus precludes its ability to model ecological interactions. However, this concern has been addressed, leading to constuctions of superprocesses that account for interactions among individuals. In the next subsection we summarize the main results in this area and introduce the SPDE that provides the basis for our approach to theoretical evolutionary ecology.

@@

With solutions to SPDE (\ref{SPDE}) well defined for any growth rate bounded above, we can calculate the total mass process $N(t)$ using the mild solution of (\ref{SPDE}) with $f(x)\equiv1\in C_b^2(\mathbb{R})$ (the symbol "$\equiv$" means equal to for every $x$). That is,

\begin{multline}
N(t)=N(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)\cdot1+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}1\right)+1\sqrt{V\nu(x,s)}\dot W(x,s)dsdx \\
=N(0)+\int_0^t\bar m(s)N(s)dt+\int_0^t\sqrt{VN(s)}d\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)}),
\end{multline}
where
\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}
and
\begin{equation}
\int_0^td\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)})=\int_0^t\int_\mathbb{R}\frac{\sqrt{\nu(x,s)}}{\sqrt{\int_\mathbb{R}\nu(x,s)dx}}\dot W(x,s)dxds.
\end{equation}
Setting $W_1(t)=\hat{\mathbf{W}}_t(\sqrt{\nu(x,t)})$, we can use traditional stochastic differential notation to write
\begin{equation}
dN=\bar mNdt+\sqrt{VN}dW_1.
\end{equation}
To find the associated SDE for $\bar x(t)$ and $\sigma^2(t)$, we want to repeat the same approach for $f(x)=x,x^2$ and apply It&ocirc;'s lemma. However, for these cases $f\notin C^2_b(\mathbb{R})$ since $f$ will not be bounded. However, if we can show that $\int_\mathbb{R}(|x|+x^2+x^4)\nu(x,t)dx<+\infty$ for all $t>0$ given this condition is satisfied by $\nu(x,0)$, then we can apply the mild solution of (\ref{SPDE}) to derive SDE for $\bar x(t)$ and $\sigma^2(t)$. To illustrate, let us suppose this is the case. Setting $\tilde x(t)=\int_\mathbb{R}x\nu(x,t)dx$, we have
\begin{equation}
\tilde x(t)=\tilde x(0)+\int_0^t\int_\mathbb{R}\nu(x,s)m(\nu,x)x+x\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Similarly, setting $\tilde{\tilde\sigma}^2(t)=\int_\mathbb{R}x^2\nu(x,t)dx$, we have
\begin{equation}
\tilde{\tilde\sigma}^2(t)=\tilde{\tilde\sigma}^2(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)x^2+\mu\right)+x^2\sqrt{V\nu(x,s)}\dot W(x,s)dxds.
\end{equation}
Since $\bar x(t)=\tilde x(t)/N(t)$ and $\sigma^2(t)=\tilde{\tilde\sigma}^2(t)/N(t)-\bar x^2(t)$, we can use It&ocirc;'s lemma to derive SDE for $\bar x(t)$ and $\sigma^2(t)$, which we perform in SM \S\ref{SDE_DERIV}. We do not make any general assertions about the existence or uniqueness of $\bar x(t)$ or $\sigma^2(t)$.
